---
title: "Correlation visualization capabilities using **plotly** in R"
author: 
  - name: "Mykhailo Andrusiak"
  - name: "Harmanjoт Singh"
date: "05 January 2025"
format:
  html:
    navigation: true
toc: true
toc_depth: 2
---

# Creating an Interactive Correlation Plot with Filtering Capabilities

## Research Question

**What are we analyzing?**  
We aim to create an interactive correlation plot using `plotly` that allows filtering to observe how correlations change across different groups (e.g., species in the Iris dataset).

---

### Step 1: Loading Libraries and Data

Loads the required libraries and the Iris dataset for analysis.
```{r, message=FALSE, warning=FALSE}
library(plotly)
library(dplyr)

data(iris)
head(iris)
```

### Step 2: Defining Regression and Correlation Line Function

Creates a custom function to calculate regression lines and correlation coefficients for subsets of data (based on species

```{r}
regression_lines <- function(data, species) {
  data_sp <- data %>% filter(Species == species)
  model <- lm(Petal.Length ~ Sepal.Length, data = data_sp) 
  
  x_range <- seq(min(data_sp$Sepal.Length), max(data_sp$Sepal.Length), length.out = 100)
  y_range <- predict(model, newdata = data.frame(Sepal.Length = x_range))
  
  cor_value <- round(cor(data_sp$Sepal.Length,
                         data_sp$Petal.Length), 2) # Calculate correlation coefficient
  
  data.frame(Sepal.Length = x_range, Petal.Length = y_range, Species = species, Correlation = cor_value)
}
```

### Step 3: Generating Regression Lines and Correlation Coefficients

Generates regression lines and calculates correlation coefficients for each species (Setosa, Versicolor, Virginica).

```{r}
lines_setosa <- regression_lines(iris, "setosa")
lines_versicolor <- regression_lines(iris, "versicolor")
lines_virginica <- regression_lines(iris, "virginica")
```

### Step 4: Creating the Interactive Plot with Data Points

Creates a scatter plot of Sepal.Length vs Petal.Length and includes a filter for species.

```{r}
fig <- plot_ly(
  data = iris,                                              # Data source
  x = ~Sepal.Length,                                       # X-axis: Sepal Length
  y = ~Petal.Length,                                       # Y-axis: Petal Length
  color = ~Species,                                        # Color by Species
  colors = c('blue', 'orange', 'green'),                   # Define colors for each species
  type = 'scatter',                                        # Specify the plot type as scatter
  mode = 'markers',                                        # Display markers
  transforms = list(
    list(
      type = 'filter',                                      # Add a filter transform
      target = ~Species,                                    # Target the Species variable
      operation = '=',                                      # Operation type: equals
      value = "setosa"                                      # Initial filter value: setosa
    )
  )
)
```

### Step 5: Adding Regression Lines to the Plot

Adds dynamic regression lines for each species to the scatter plot.
```{r}
fig <- fig %>%
  add_lines(
    data = lines_setosa,                                      # Data for Setosa regression line
    x = ~Sepal.Length,                                       # X values for the line
    y = ~Petal.Length,                                       # Y values for the line
    line = list(color = 'blue', width = 1),                  # Line style
    name = paste("Setosa (r =", lines_setosa$Correlation[1], ")")  # Legend name with correlation
  ) %>%
  add_lines(
    data = lines_versicolor,                                  # Data for Versicolor regression line
    x = ~Sepal.Length,                                       # X values for the line
    y = ~Petal.Length,                                       # Y values for the line
    line = list(color = 'orange', width = 1),                # Line style
    name = paste("Versicolor (r =", lines_versicolor$Correlation[1], ")")  # Legend name with correlation
  ) %>%
  add_lines(
    data = lines_virginica,                                   # Data for Virginica regression line
    x = ~Sepal.Length,                                       # X values for the line
    y = ~Petal.Length,                                       # Y values for the line
    line = list(color = 'green', width = 1),                 # Line style
    name = paste("Virginica (r =", lines_virginica$Correlation[1], ")")  # Legend name with correlation
  )
```

### Step 6: Adding Filtering Capabilities to the Plot

Includes a dropdown menu to filter the plot by species or show all data points together.

```{r}
fig <- fig %>%
  layout(
    title = "Dynamic Lines with Correlation Coefficients",          # Plot title
    xaxis = list(title = "Sepal Length (cm)"),                      # X-axis title
    yaxis = list(title = "Petal Length (cm)"),                      # Y-axis title
    updatemenus = list(
      list(
        buttons = list(
          list(
            method = "restyle",                                     # Method to update the plot
            args = list("transforms[0].value", "setosa"),          # Filter for Setosa
            label = "Iris-setosa"                                   # Button label
          ),
          list(
            method = "restyle",                                     # Method to update the plot
            args = list("transforms[0].value", "versicolor"),      # Filter for Versicolor
            label = "Iris-versicolor"                               # Button label
          ),
          list(
            method = "restyle",                                     # Method to update the plot
            args = list("transforms[0].value", "virginica"),       # Filter for Virginica
            label = "Iris-virginica"                                # Button label
          ),
          list(
            method = "restyle",                                     # Method to update the plot
            args = list("transforms[0].value", unique(iris$Species)),  # Show all species
            label = "All"                                           # Button label
          )
        ),
        direction = "down",                                           # Dropdown direction
        x = 0.1,                                                      # X position of the dropdown
        y = 1.15,                                                     # Y position of the dropdown
        showactive = TRUE                                             # Show active button
      )
    )
  )

fig
```

### About the Plot

**What the plot shows:**  
This interactive plot visualizes the relationship between `Sepal.Length` and `Petal.Length` for the Iris dataset, with regression lines and correlation coefficients for each species. Key features include:  

1. **Filtering:**  
   - Use the dropdown menu to filter the data by species (`Setosa`, `Versicolor`, or `Virginica`) or view all species together.
   
2. **Dynamic Regression Lines:**  
   - Each species has its regression line displayed with its correlation coefficient (`r`). These lines dynamically adjust based on the selected filter.

3. **Interactive Legend:**  
   - You can click on the legend items to toggle the visibility of specific data points or regression lines, allowing for customized views.

4. **Scalability:**  
   - The plot supports zooming and panning, making it easy to focus on specific data points or regions of interest.  
   - Use the built-in toolbar to reset the view, save the graph, or activate zoom modes.

This plot is an excellent tool for exploring both individual and combined group correlations, enabling deeper insights into the data structure.

---

## Conclusion

**Key Insights:**

1. **Interactive Filtering and Group Analysis:**  
   - The dropdown menu allows users to isolate specific species or compare all groups together, providing flexibility in exploring correlations.

2. **Dynamic Regression Lines and Correlations:**  
   - The regression lines, annotated with `r` values, provide a clear understanding of the strength and direction of linear relationships for each group.  
   - This makes it easy to compare correlations across different subsets.

3. **Legend Interactions and Zooming:**  
   - The interactive legend enables users to toggle data visibility, simplifying the exploration of specific subsets or regression lines.  
   - Zooming and panning features improve usability, especially when analyzing dense or overlapping data points.

**Recommendation:**  
This approach is ideal for datasets with well-defined groups (e.g., categories or classes). Use dynamic filtering to explore correlations efficiently across subsets. The interactive legend and zoom features further enhance the user experience, making it suitable for exploratory data analysis and presentations to diverse audiences.

# Correlation Visualization in Small and Large Datasets

## Research Question

**What are we analyzing?**  
We aim to determine which correlation visualization methods are most effective for small and large datasets.  
We will use **Heatmaps** for a quick overview of correlations and **Scatter Matrix** plots for a detailed examination of relationships between variables.

---

### Step 1: Importing Libraries and Preparing the Data

•	Imports additional library for loading data
```{r,message=FALSE, warning=FALSE}
library(ggplot2)  # For diamonds
```
### Step 2: Small Dataset: Correlation Heatmap

•	Loads the mtcars dataset.  
•	Calculates the correlation matrix for all numeric variables.
```{r}
data("mtcars")
small_data <- mtcars

small_corr <- round(cor(small_data), 2)
```
•	Creates a Heatmap to visualize the correlations.
```{r}
fig1 <- plot_ly(
  data = small_data,                              # Data source
  x = colnames(small_corr),                       # X-axis: variable names
  y = colnames(small_corr),                       # Y-axis: variable names
  z = small_corr,                                 # Z-axis: correlation values
  type = "heatmap",                               # Specify the type as heatmap
  colorscale = "Viridis",                         # Color scale for the heatmap
  text = round(small_corr, 2),                    # Text to display on hover
  hoverinfo = "x+y+text"                           # Information to display on hover
) %>%
  layout(
    title = "Heatmap of Correlation (Small Dataset: mtcars)",  # Title of the plot
    xaxis = list(title = "Variables"),                         # X-axis title
    yaxis = list(title = "Variables"),                         # Y-axis title
    annotations = list(
      x = rep(colnames(small_corr), each = nrow(small_corr)), # X positions for annotations
      y = rep(colnames(small_corr), ncol(small_corr)),        # Y positions for annotations
      text = as.character(round(small_corr, 2)),              # Text for annotations
      showarrow = FALSE,                                       # Hide arrows in annotations
      font = list(size = 12, color = "white")                 # Font size and color for annotations
    )
  )
fig1
```
**About the plot:**  
The Heatmap displays the correlations between numeric variables in the mtcars dataset.  
•	Yellow color: strong positive correlations.  
•	Purple color: strong negative correlations.  
This allows for a quick identification of the strongest and weakest relationships.  

### Step 3: Small Dataset: Scatter Matrix  
Generates a Scatter Matrix for key variables mpg, hp, wt, and qsec in the mtcars dataset.
```{r}
fig2 <- plot_ly(
  data = small_data,                             # Data source
  type = "splom",                                # Specify the type as scatter plot matrix
  dimensions = list(
    list(label = "mpg", values = ~mpg),          # Define the first dimension
    list(label = "hp", values = ~hp),            # Define the second dimension
    list(label = "wt", values = ~wt),            # Define the third dimension
    list(label = "qsec", values = ~qsec)         # Define the fourth dimension
  )
) %>%
  layout(
    title = "Scatter Matrix (Small Dataset: mtcars)"  # Title of the plot
  )

fig2
```
**About the plot:**  
The Scatter Matrix visualizes pairwise relationships between key variables in the mtcars dataset, along with their distributions. For example, mpg shows a strong negative correlation with hp and wt.  

### Step 4: Large Dataset: Correlation Heatmap  
•	Samples 1,000 rows from the diamonds dataset.  
•	Computes the correlation matrix for numeric variables.  
```{r}
data(diamonds)
large_data <- diamonds %>% sample_n(1000)  

large_corr <- large_data %>% 
  select_if(is.numeric) %>% 
  cor() %>% 
  round(2)
```
•	Creates a Heatmap to visualize the correlations.
```{r}
fig3 <- plot_ly(
  x = colnames(large_corr),                          # X-axis: variable names
  y = colnames(large_corr),                          # Y-axis: variable names
  z = large_corr,                                    # Z-axis: correlation values
  type = "heatmap",                                  # Specify the type as heatmap
  colorscale = "Viridis",                            # Color scale for the heatmap
  text = round(large_corr, 2),                       # Text to display on hover
  hoverinfo = "x+y+text"                              # Information to display on hover
) %>%
  layout(
    title = "Heatmap of Correlation (Large Dataset: diamonds)",  # Title of the plot
    xaxis = list(title = "Variables"),                           # X-axis title
    yaxis = list(title = "Variables"),                           # Y-axis title
    annotations = list(
      x = rep(colnames(large_corr), each = nrow(large_corr)),    # X positions for annotations
      y = rep(colnames(large_corr), ncol(large_corr)),           # Y positions for annotations
      text = as.character(round(large_corr, 2)),                 # Text for annotations
      showarrow = FALSE,                                         # Hide arrows in annotations
      font = list(size = 12, color = "white")                   # Font size and color for annotations
    )
  )

fig3
```
**About the plot:**  
The Heatmap shows correlations between numeric variables in the diamonds subset. Strong positive correlations are visible between carat and size-related variables (x, y, z), highlighted in yellow.  

### Step 5: Large Dataset: Scatter Matrix  
•	Generates a Scatter Matrix for all numeric variables in the diamonds dataset sample.
```{r}
numeric_data <- large_data[sapply(large_data, is.numeric)]

fig4 <- plot_ly(
  data = numeric_data,                              # Data source
  type = "splom",                                   # Specify the type as scatter plot matrix
  dimensions = lapply(names(numeric_data), function(col) {
    list(label = col, values = numeric_data[[col]])  # Define each dimension dynamically
  })
) %>%
  layout(
    title = "Scatter Matrix (Large Dataset: diamonds)",  # Title of the plot
    margin = list(b = 50)                                 # Adjust bottom margin
  )

fig4
```
**About the plot:**  
The Scatter Matrix for the diamonds dataset sample shows pairwise relationships between numeric variables. For instance, carat has a clear positive linear relationship with x, y, and z.

## Conclusion

**Key Findings:**  
  **1.	Heatmap:**  
    •	Effective for quickly assessing correlations in both small and large datasets.  
    •	Color gradients make it easy to identify the strongest and weakest relationships.  
  **2.	Scatter Matrix:**  
    •	More informative for detailed pairwise analysis of variables.  
    •	Suitable for small datasets or selected subsets of variables in large datasets.  
    
# Interactive Heatmaps for Correlation Visualization

## Research Question
This analysis demonstrates how to create an interactive heatmap of correlations to visualize relationships between numerical variables. The goal is to explore correlation coefficients and observe how they change when the input data is modified. We will use the `diamonds` dataset from the `ggplot2` package and visualize relationships using `plotly` for interactivity.

### Step 1: Select Numerical Variables

Since the necessary libraries have already been loaded we identify and select only the numerical variables from the dataset to compute correlations. This ensures the analysis focuses solely on numerical relationships.

```{r}
# Identify numerical columns in the dataset
numeric_columns <- sapply(diamonds, is.numeric)

# Create a new dataframe with only numerical variables
diamonds_numeric <- diamonds[, numeric_columns]
```

The `sapply()` function checks each column to determine if it is numeric. We use this information to subset the dataset and create a new dataframe `diamonds_numeric` containing only numerical variables.

### Step 2: Calculate the Correlation Matrix

Next, we calculate the correlation matrix for the selected numerical variables. The correlation matrix quantifies the pairwise relationships between variables, ranging from -1 (perfect negative correlation) to +1 (perfect positive correlation).

```{r}
# Compute the correlation matrix
correlation_matrix <- cor(diamonds_numeric, use = "complete.obs")
```

The `cor()` function computes pairwise correlations. Setting `use = "complete.obs"` ensures rows with missing values are excluded.

### Step 3: Create an Interactive Correlation Heatmap

To visualize the correlation matrix interactively, we define a function that generates a heatmap using `plotly`. This heatmap allows users to explore correlations interactively by hovering over the cells.

```{r}
# Create a simple function to generate an interactive heatmap with numbers
create_correlation_heatmap <- function(correlation_matrix) {
  plot_ly(
    x = colnames(correlation_matrix),   # Variable names for the x-axis
    y = rownames(correlation_matrix),   # Variable names for the y-axis
    z = correlation_matrix,             # Correlation coefficients as z-values
    type = "heatmap",                   # Specify that we want a heatmap
    colorscale = "RdBu",                # Use a red-blue color scale
    reversescale = TRUE,                # Reverse the color scale for red-positive correlations
    text = round(correlation_matrix, 2), # Add rounded correlation values as text
    texttemplate = "%{text}"            # Display numbers on the heatmap
  ) %>%
    layout(
      title = "Interactive Correlation Heatmap",
      xaxis = list(title = "Variables"),
      yaxis = list(title = "Variables")
    )
}

# Call the function to display the heatmap
create_correlation_heatmap(correlation_matrix)
```

The function takes a correlation matrix as input and:

1. Maps variables to the x and y axes.

2. Uses a diverging color scale (`RdBu`) to highlight both positive (blue) and negative (red) correlations.

3. Displays rounded correlation values on the heatmap.

4. Creates an interactive visualization where users can hover over cells to view details.

### Step 4: Modify the Dataset and Observe Changes

To analyze how correlations change with modified input data, we create a filtered version of the dataset that includes only diamonds with `carat` values less than 2. This focuses on a subset of diamonds and recalculates the correlation matrix.

```{r}
# Filter the dataset to include only diamonds with carat < 2
diamonds_filtered <- diamonds_numeric %>% filter(carat < 2)

# Compute the correlation matrix for the filtered dataset
correlation_matrix_filtered <- cor(diamonds_filtered, use = "complete.obs")

# Generate a heatmap for the filtered dataset
create_correlation_heatmap(correlation_matrix_filtered)
```

Here:

1. The `filter()` function removes rows where `carat` is 2 or more.

2. A new correlation matrix is calculated using the filtered data.

3. A heatmap is generated for the modified dataset, allowing for comparison with the original heatmap.

## Observations

1. Correlations can shift when the input data is modified. For example, restricting the range of `carat` may reduce the variability in related variables, potentially weakening their correlations.

2. By comparing the heatmaps, you can visually identify which relationships remain strong and which weaken in the filtered dataset.

## Conclusion

Interactive heatmaps are a powerful tool for exploring and visualizing correlations between numerical variables. They allow users to engage with the data dynamically and observe how changes in input data affect relationships. This approach makes complex data more accessible and interpretable, particularly for exploratory data analysis.

# Analyzing the Impact of Color Scales on Correlation Heatmaps

## Research Question

This document analyzes how different color scales (sequential and diverging) affect the perception of correlation strength in heatmaps. Using the `diamonds` dataset from the `ggplot2` package, we will:

1. Create two correlation heatmaps separately:
   - One with a **diverging color scale** (`RdBu`) to emphasize both the direction (positive/negative) and magnitude of correlations.
   - Another with a **sequential color scale** (`Viridis`) to emphasize the strength of correlations, regardless of their direction.
2. Compare these visualizations to identify their strengths and weaknesses.
3. Determine the top 5 strongest positive and negative correlations in the dataset.

---

### Step 1: Create a Heatmap with a Diverging Color Scale

Since the necessary libraries have already been loaded, the diamonds dataset has been filtered and Correlation Matrix has been calculated. Next, we visualize the correlation matrix using a diverging color scale (`RdBu`). Diverging color scales are particularly useful for highlighting the direction of correlations. Strong positive correlations appear in one color (e.g., red), strong negative correlations appear in another (e.g., blue), and neutral correlations are displayed as white.

```{r}
# Create a heatmap with a diverging color scale (RdBu)
plot_ly(
  x = colnames(correlation_matrix),          # X-axis: variable names
  y = rownames(correlation_matrix),          # Y-axis: variable names
  z = correlation_matrix,                    # Z-axis: correlation values
  type = "heatmap",                         # Heatmap type
  colorscale = "RdBu",                      # Diverging color scale
  text = round(correlation_matrix, 2),       # Display rounded correlation values
  texttemplate = "%{text}"                 # Text formatting
) %>%
  layout(
    title = "Heatmap with Diverging Color Scale (RdBu)",
    xaxis = list(title = "Variables"),
    yaxis = list(title = "Variables")
  )
```

This block creates an interactive heatmap using the `plot_ly` function. The `colorscale = "RdBu"` argument applies a diverging color scheme, where red and blue emphasize positive and negative correlations, respectively. Neutral correlations are represented by white, making them easily distinguishable. Rounded correlation values are displayed on the heatmap for clarity.

---

### Step 2: Create a Heatmap with a Sequential Color Scale

We now create another heatmap, this time using a sequential color scale (`Viridis`). Sequential color scales emphasize the strength of correlations without differentiating between positive and negative values.

```{r}
# Create a heatmap with a sequential color scale (Viridis)
plot_ly(
  x = colnames(correlation_matrix),          # X-axis: variable names
  y = rownames(correlation_matrix),          # Y-axis: variable names
  z = correlation_matrix,                    # Z-axis: correlation values
  type = "heatmap",                         # Heatmap type
  colorscale = "Viridis",                   # Sequential color scale
  text = round(correlation_matrix, 2),       # Display rounded correlation values
  texttemplate = "%{text}"                 # Text formatting
) %>%
  layout(
    title = "Heatmap with Sequential Color Scale (Viridis)",
    xaxis = list(title = "Variables"),
    yaxis = list(title = "Variables")
  )
```

This code uses the same approach as before but applies a sequential color scale (`Viridis`). In this scale, darker colors represent lower correlation values, and brighter colors represent higher values. It is particularly effective for highlighting the magnitude of relationships but does not indicate their direction.

---

### Step 3: Identify Top Positive and Negative Correlations

To complement the visual analysis, we identify the strongest positive and negative correlations in the dataset. This involves reshaping the correlation matrix into a long format and sorting the values to find the top correlations.

```{r}
# Convert the correlation matrix into a data frame for sorting
correlation_df <- as.data.frame(as.table(correlation_matrix)) %>%
  filter(Var1 != Var2) %>%            # Exclude diagonal (self-correlations)
  arrange(desc(Freq))                 # Sort by correlation values

# Top 5 positive correlations
top_positive <- correlation_df %>% filter(Freq > 0) %>% head(5)

# Top 5 negative correlations
top_negative <- correlation_df %>% filter(Freq < 0) %>% tail(5)

# Display results
print("Top 5 Positive Correlations:")
print(top_positive)

print("Top 5 Negative Correlations:")
print(top_negative)
```

In this step, we convert the correlation matrix into a long-format data frame using `as.table`. This allows us to filter out self-correlations (diagonal elements where variables are correlated with themselves) and sort the correlations in descending order. We then extract the top 5 strongest positive and negative correlations for further interpretation.

---

## Analysis and Conclusion

This analysis highlights the impact of different color scales on the interpretation of correlation heatmaps. The diverging color scale (`RdBu`) effectively distinguishes between positive and negative correlations, making it ideal for identifying the direction of relationships. However, it may be visually overwhelming due to its strong contrasts. In contrast, the sequential color scale (`Viridis`) provides a smoother gradient that emphasizes the strength of correlations but does not differentiate their direction. 

The choice of color scale depends on the goals of the analysis. Use diverging scales when the direction of correlations is important and sequential scales when focusing solely on the magnitude of relationships.

