---
output: html_document
---

# Correlation Visualization in Small and Large Datasets

## Research Question

**What are we analyzing?**  
We aim to determine which correlation visualization methods are most effective for small and large datasets.  
We will use **Heatmaps** for a quick overview of correlations and **Scatter Matrix** plots for a detailed examination of relationships between variables.

---

## Step 1: Importing Libraries and Preparing the Data

•	Imports the necessary libraries for working with data and visualizations.
```{r,message=FALSE, warning=FALSE}
library(plotly)
library(ggplot2)  # For diamonds
library(dplyr)
```
## Step 2: Small Dataset: Correlation Heatmap

•	Loads the mtcars dataset.  
•	Calculates the correlation matrix for all numeric variables.
```{r}
data("mtcars")
small_data <- mtcars

small_corr <- round(cor(small_data), 2)
```
•	Creates a Heatmap to visualize the correlations.
```{r}
fig1 <- plot_ly(
  data = small_data,                              # Data source
  x = colnames(small_corr),                       # X-axis: variable names
  y = colnames(small_corr),                       # Y-axis: variable names
  z = small_corr,                                 # Z-axis: correlation values
  type = "heatmap",                               # Specify the type as heatmap
  colorscale = "Viridis",                         # Color scale for the heatmap
  text = round(small_corr, 2),                    # Text to display on hover
  hoverinfo = "x+y+text"                           # Information to display on hover
) %>%
  layout(
    title = "Heatmap of Correlation (Small Dataset: mtcars)",  # Title of the plot
    xaxis = list(title = "Variables"),                         # X-axis title
    yaxis = list(title = "Variables"),                         # Y-axis title
    annotations = list(
      x = rep(colnames(small_corr), each = nrow(small_corr)), # X positions for annotations
      y = rep(colnames(small_corr), ncol(small_corr)),        # Y positions for annotations
      text = as.character(round(small_corr, 2)),              # Text for annotations
      showarrow = FALSE,                                       # Hide arrows in annotations
      font = list(size = 12, color = "white")                 # Font size and color for annotations
    )
  )
fig1
```
**About the plot:**  
The Heatmap displays the correlations between numeric variables in the mtcars dataset.  
•	Yellow color: strong positive correlations.  
•	Purple color: strong negative correlations.  
This allows for a quick identification of the strongest and weakest relationships.  

## Step 3: Small Dataset: Scatter Matrix  
Generates a Scatter Matrix for key variables mpg, hp, wt, and qsec in the mtcars dataset.
```{r}
fig2 <- plot_ly(
  data = small_data,                             # Data source
  type = "splom",                                # Specify the type as scatter plot matrix
  dimensions = list(
    list(label = "mpg", values = ~mpg),          # Define the first dimension
    list(label = "hp", values = ~hp),            # Define the second dimension
    list(label = "wt", values = ~wt),            # Define the third dimension
    list(label = "qsec", values = ~qsec)         # Define the fourth dimension
  )
) %>%
  layout(
    title = "Scatter Matrix (Small Dataset: mtcars)"  # Title of the plot
  )

fig2
```
**About the plot:**  
The Scatter Matrix visualizes pairwise relationships between key variables in the mtcars dataset, along with their distributions. For example, mpg shows a strong negative correlation with hp and wt.  

## Step 4: Large Dataset: Correlation Heatmap  
•	Samples 1,000 rows from the diamonds dataset.  
•	Computes the correlation matrix for numeric variables.  
```{r}
data("diamonds")
large_data <- diamonds %>% sample_n(1000)  

large_corr <- large_data %>% 
  select_if(is.numeric) %>% 
  cor() %>% 
  round(2)
```
•	Creates a Heatmap to visualize the correlations.
```{r}
fig3 <- plot_ly(
  x = colnames(large_corr),                          # X-axis: variable names
  y = colnames(large_corr),                          # Y-axis: variable names
  z = large_corr,                                    # Z-axis: correlation values
  type = "heatmap",                                  # Specify the type as heatmap
  colorscale = "Viridis",                            # Color scale for the heatmap
  text = round(large_corr, 2),                       # Text to display on hover
  hoverinfo = "x+y+text"                              # Information to display on hover
) %>%
  layout(
    title = "Heatmap of Correlation (Large Dataset: diamonds)",  # Title of the plot
    xaxis = list(title = "Variables"),                           # X-axis title
    yaxis = list(title = "Variables"),                           # Y-axis title
    annotations = list(
      x = rep(colnames(large_corr), each = nrow(large_corr)),    # X positions for annotations
      y = rep(colnames(large_corr), ncol(large_corr)),           # Y positions for annotations
      text = as.character(round(large_corr, 2)),                 # Text for annotations
      showarrow = FALSE,                                         # Hide arrows in annotations
      font = list(size = 12, color = "white")                   # Font size and color for annotations
    )
  )

fig3
```
**About the plot:**  
The Heatmap shows correlations between numeric variables in the diamonds subset. Strong positive correlations are visible between carat and size-related variables (x, y, z), highlighted in yellow.  

## Step 5: Large Dataset: Scatter Matrix  
•	Generates a Scatter Matrix for all numeric variables in the diamonds dataset sample.
```{r}
numeric_data <- large_data[sapply(large_data, is.numeric)]

fig4 <- plot_ly(
  data = numeric_data,                              # Data source
  type = "splom",                                   # Specify the type as scatter plot matrix
  dimensions = lapply(names(numeric_data), function(col) {
    list(label = col, values = numeric_data[[col]])  # Define each dimension dynamically
  })
) %>%
  layout(
    title = "Scatter Matrix (Large Dataset: diamonds)",  # Title of the plot
    margin = list(b = 50)                                 # Adjust bottom margin
  )

fig4
```
**About the plot:**  
The Scatter Matrix for the diamonds dataset sample shows pairwise relationships between numeric variables. For instance, carat has a clear positive linear relationship with x, y, and z.

## Conclusion

**Key Findings:**  
  **1.	Heatmap:**  
    •	Effective for quickly assessing correlations in both small and large datasets.  
    •	Color gradients make it easy to identify the strongest and weakest relationships.  
  **2.	Scatter Matrix:**  
    •	More informative for detailed pairwise analysis of variables.  
    •	Suitable for small datasets or selected subsets of variables in large datasets.  